Quantum cluster approach to the topological invariants in correlated Chern insulators

### 2019-05-20 paper/paper.tex
1st reply to NJP referees.

\par Band insulators with nontrivial topology have become one of the central interests in condensed matter physics since the discovery of topological insulators \cite{HK_RMP2010,QZ_RMP2011}. Beyond the traditional Landau paradigm, topological insulators cannot be distinguished from ordinary ones by local order parameters, but are featured by gapless edges states on open boundaries \cite{H_PRL1988,K_PU2001,KM_PRL2005,BHZ_S2006,LYGJ_PRB2016,LGLW_NJP2017}. The existence of such edge modes is linked to the emergence of nonzero topological invariants of the bulk energy bands via the bulk-edge correspondence \cite{QWZ_PRB2006}. Thus the identification of a topological invariant is of particular importance to characterize different phases of band insulators. Indeed, in terms of such topological invariants constructed from the free Hamiltonians, a complete classification has been developed for non-interacting systems in different symmetry classes \cite{SRFL_PRB2008}.

\par The presence of strong interactions between electrons enriches the physics of topological bands, and brings many challenges to this area at the same time. One of such demanding questions is how one can compute the topological invariant in an interacting system. A pioneering work by Volovik \cite{V2003} expresses the Chern number \cite{TKNN_PRL1982} of a two dimensional Chern insulator in terms of the single particle Green's functions. This formula is ready to be generalized to deal with correlated Chern insulators by a simple substitution of the free propagator with the interacting one. A great simplification of this formula was proposed in Ref. \cite{WZ_PRX2012} that the zero-energy single particle Green's function alone is sufficient to calculate the interacting Chern number via the Kubo formula. Later, the concept of the so-called topological Hamiltonian \cite{WY_JPCM2013}, i.e. the minus inverse of the dressed single particle Green's function, was shown to apply equally in other systems with diverse topological classifications \cite{WA_PRB2013,LEGW_PRB2013,GMCBB_NJP2015,LS_PRB2018,SK_PRB2018}, such as the quantum spin Hall insulators \cite{LEGW_PRB2013,GMCBB_NJP2015}, topological superconductors \cite{LS_PRB2018}, interacting Su-Schrieffer-Heeger chains \cite{SK_PRB2018}, etc.

\par Thus, the success of the topological Hamiltonian relies on a credible computation of the interacting single particle Green's functions. In recent years a variety of quantum many-body algorithms gathering under the name of quantum cluster approaches \cite{MJPH_RMP2005,S_arXiv2008} have been developed to calculate such single particle Green's functions of a broad range of strongly correlated fermionic systems \cite{SPP_PRL2000,HMJK_PRB2000,KSPB_PRL2001,PAD_PRL2003,ST_PRL2004,BKSTP_EPL2009,KYXL_PRB2011,YXL_PRL2011,YL_PRB2012,LYXL_PRB2014} and bosonic systems \cite{KD_JPCM2006,KAL_PRB2011,YWDYL_PRB2018}. They share the basic idea of solving a reference system composed of isolated finite-sized clusters that are sometimes embedded in a uncorrelated bath media. They give access to nontrivial many body effects and can be unified within the variational principle based on the self energy functional approach \cite{P_EPJB2003}. These methods include cluster perturbation theory (CPT) \cite{SPP_PRL2000}, dynamical cluster approach (DCA) \cite{HMJK_PRB2000}, cellular dynamical mean field theory (CDMFT) \cite{KSPB_PRL2001}, variational cluster approach (VCA) \cite{PAD_PRL2003}, cluster dynamical impurity approximation (CDIA) \cite{BKSTP_EPL2009}, etc, and have been widely applied to detect the topological properties of interacting topological band insulators \cite{YXL_PRL2011,GMCBB_NJP2015,WFSM_PRB2016,LS_PRB2018}.

@@

@@

\par We want to remark that the actual phase diagram of the Haldane Hubbard model in the moderate interacting regime has been highly debated over the past several years, namely different methods predicted quite different results \cite{HZKL_PRB2011-1,HZKL_PRB2011-2,MR_PRB2013,HRP_PRB2015,ZSWZ_PRB2015,HCPP_PRL2016,VSLTHT_PRL2016,IWT_PRB2016,ASHP_PRB2016,GJMP_PRB2016,GR_NJP2018,LTTNNH_PBCM2018}. In particular, even whether the intermediate phases exist has not come to an agreement, let alone their physical properties. Due to the limited power of quantum cluster approaches, we have no intention to settle down this difficult problem in this paper. However, it is safe to claim that in the framework of CPT/VCA, the intermediate phase (NMI) exists and it is not topological in the sense that it has no nonzero interacting bulk Chern number nor gapless states or zero-energy zeros of the single particle Green's function on edges. Besides, the spinless free Haldane model has been realized in recent cold atom experiments using the shaking lattice technique \cite{JMDRLUGE_N2014}. We hope our claim can be tested in future experiments.

>>>>

\par Band insulators with nontrivial topology have become one of the central interests in condensed matter physics since the discovery of topological insulators \cite{HK_RMP2010,QZ_RMP2011}. Beyond the traditional Landau paradigm, topological insulators cannot be distinguished from ordinary ones by local order parameters, but are featured by gapless edges states on open boundaries \cite{H_PRL1988,K_PU2001,KM_PRL2005,BHZ_S2006,LYGJ_PRB2016,LGLW_NJP2017}. The existence of such edge modes is linked to the emergence of nonzero topological invariants of the bulk energy bands via the bulk-edge correspondence \cite{QWZ_PRB2006}. Thus the identification of a topological invariant is of particular importance to characterize different phases of band insulators. Indeed, in terms of such topological invariants constructed from the free Hamiltonians, a complete classification has been developed for non-interacting systems in different symmetry classes \cite{SRFL_PRB2008,CTSR_RMP2016}.

\par The presence of strong interactions between electrons enriches the physics of topological bands, and brings many challenges to this area at the same time \cite{HA_JPCM2013,R_RPP2018}. One of such demanding questions is how one can compute the topological invariant in an interacting system. A pioneering work by Volovik \cite{V2003} expresses the Chern number \cite{TKNN_PRL1982} of a two dimensional Chern insulator in terms of the single particle Green's functions. This formula is ready to be generalized to deal with correlated Chern insulators by a simple substitution of the free propagator with the interacting one. A great simplification of this formula was proposed in Ref. \cite{WZ_PRX2012} that the zero-energy single particle Green's function alone is sufficient to calculate the interacting Chern number via the Kubo formula. Later, the concept of the so-called topological Hamiltonian \cite{WY_JPCM2013}, i.e. the minus inverse of the dressed single particle Green's function, was shown to apply equally in other systems with diverse topological classifications \cite{WA_PRB2013,LEGW_PRB2013,HWGF_PRB2013,GMCBB_NJP2015,LS_PRB2018,SK_PRB2018}, such as the quantum spin Hall insulators \cite{LEGW_PRB2013,HWGF_PRB2013,GMCBB_NJP2015}, topological superconductors \cite{LS_PRB2018}, interacting Su-Schrieffer-Heeger chains \cite{SK_PRB2018}, etc.

\par Thus, the success of the topological Hamiltonian relies on a credible computation of the interacting single particle Green's functions. In recent years a variety of quantum many-body algorithms gathering under the name of quantum cluster approaches \cite{MJPH_RMP2005,S_arXiv2008} have been developed to calculate such single particle Green's functions of a broad range of strongly correlated fermionic systems \cite{SPP_PRL2000,HMJK_PRB2000,KSPB_PRL2001,PAD_PRL2003,ST_PRL2004,SS_PRL2008,BKSTP_EPL2009,KYXL_PRB2011,YXL_PRL2011,WRLH_PRB2012,YL_PRB2012,HD_PRL2013,LRTR_PRB2014,LYXL_PRB2014,RLRT_PRL2015} and bosonic systems \cite{KD_JPCM2006,KAL_PRB2011,YWDYL_PRB2018}. They share the basic idea of solving a reference system composed of isolated finite-sized clusters that are sometimes embedded in a uncorrelated bath media. They give access to nontrivial many body effects and can be unified within the variational principle based on the self energy functional approach \cite{P_EPJB2003}. These methods include cluster perturbation theory (CPT) \cite{SPP_PRL2000}, dynamical cluster approach (DCA) \cite{HMJK_PRB2000}, cellular dynamical mean field theory (CDMFT) \cite{KSPB_PRL2001}, variational cluster approach (VCA) \cite{PAD_PRL2003}, cluster dynamical impurity approximation (CDIA) \cite{BKSTP_EPL2009}, etc, and have been widely applied to detect the topological properties of interacting topological band insulators \cite{YXL_PRL2011,WRLH_PRB2012,LRTR_PRB2014,GMCBB_NJP2015,WFSM_PRB2016,LS_PRB2018}.

@@

\par CPT/VCA can also calculate single particle edge states of two dimensional interacting systems. This is done by tiling clusters into a supercluster with a ribbon geometry, which is subject to open boundary condition along one direction but to periodic boundary condition along the other [e.g. see Fig. \ref{lattice}(c)]. Now the supercluster Green's function $G_0(\omega)$ in Eq. (\ref{SPGF}) is taken to be the direct sum of the single cluster Green's functions, and the couplings between individual clusters within this supercluster are treated as perturbations, i.e. they show up in the $V(\widetilde{\mathbf{k}})$ term in Eq. (\ref{SPGF}). Other procedures are exactly the same with those to compute the bulk properties.

@@

\par We want to remark that the actual phase diagram of the Haldane Hubbard model in the moderate interacting regime has been highly debated over the past several years, namely different methods predicted quite different results \cite{HZKL_PRB2011-1,HZKL_PRB2011-2,MR_PRB2013,HRP_PRB2015,ZSWZ_PRB2015,WFSM_PRB2016,HCPP_PRL2016,VSLTHT_PRL2016,IWT_PRB2016,ASHP_PRB2016,GJMP_PRB2016,GR_NJP2018,LTTNNH_PBCM2018}. In particular, even whether the intermediate phases exist has not come to an agreement, let alone their physical properties. Due to the limited power of quantum cluster approaches, we have no intention to settle down this difficult problem in this paper. It is only safe to claim that in the framework of CPT/VCA, there exist an intermediate phase (NMI), which is not topological in the sense that it has no nonzero interacting bulk Chern number nor gapless states or zero-energy zeros of the single particle Green's function on edges. Nevertheless, this claim cannot rule out the possibility of the exotic nonmagnetic phases with topological fractionalized excitations \cite{HZKL_PRB2011-2,MR_PRB2013,HRP_PRB2015,HCPP_PRL2016} because it is not guaranteed that the nontrivial topology of the fractionalized modes must manifest itself in the single particle Green's functions. It is also possible that the NMI phase is just an artifact of the CPT/VCA method. Indeed, as can be seen in Fig. \ref{phase}(a), the NMI phases connects to the well known fictitious ``spin liquid'' phase in the simple honeycomb Hubbard model with $t'=0$ \cite{MLWAM_N2010,SOY_SR2012,AH_PRX2013,HD_PRL2013}, which arises from the nonequivalent treatment of the intracluster hoppings and the intercluster ones of the six-site tiling\cite{LRTR_PRB2014}. Yet the experiences of the case with $t'=0$ may not be definitely generalized to that with a finite $t'$. In view of the effective spin model incorporated with finite-$U$ effects, a nonzero $t'$ term introduces next nearest neighbor spin interactions and ring-exchange spin interactions, which could frustrate magnetic order and stabilize exotic quantum disordered phases \cite{HRP_PRB2015,HCPP_PRL2016}. Besides, charge fluctuations are known to be possible to suppress magnetic order and favor nonmagnetic phases in the moderate interacting regime \cite{SS_PRL2008,YKK_PRL2009,RLRT_PRL2015}. Further sophisticated numerical analyses, such as the large scale density matrix renormalization group simulations, are needed to determine the phases in the moderate interacting regime.

\par However, whether the NMI phase exists has nothing to do with the reported fault here caused by the standard periodization procedure that leads to a false interacting Chern number. In fact, as mentioned in Sec. \ref{MM}, the density of states and the grand potential are not affected by the approximation employed by the periodization procedure, thus, the single particle gap $\Delta$ and the antiferromagnetic moment $M$ are irrelevant to this approximation. Note that $\Delta$ and $M$ completely determines the phase diagram shown in Fig. \ref{phase}(a). Therefore, the cause that accounts for a false interacting topological Chern number is different from the one that may result in a possible fictitious NMI. Thus, the validity of the modified computing scheme of applying quantum cluster approaches to the calculation of interacting topological invariant is independent of the existence of this specific ``NMI'' phase. In fact, it applies to all insulating phases that could either be nonmagnetic or host magnetic orders. Indeed, based on the same computing scheme, the whole AFM phase is shown to be non-topological, which invalidates the claim in a previous quantum cluster study \cite{WFSM_PRB2016} based on the false natural-unit-cell Chern number that there exists a topological region in the AFM phase in the moderate interacting regime. Besides, the noninteracting spinless Haldane model has been realized in recent cold atom experiments using the shaking lattice technique \cite{JMDRLUGE_N2014}. We hope that our claim can be tested in future experiments.
